### 2024-06-30 causal-inference.qmd
20240630 - duplicate figure label

A key tool when describing a research question or hypothesis is to create a conceptual depiction of the hypothesized causal processes.
A causal diagram depicts the hypothesized causal processes that link two or more variables.
A common form of causal diagrams is the directed acyclic graph (DAG).
DAGs provide a helpful tool to communicate about causal questions and help identify how to avoid bias (i.e., over-estimation) in associations between variables due to confounding (i.e., common causes) [@Digitale2022].
You can use the `R` package `dagitty` [@Textor2017] or the associated browser-based extension, DAGitty: https://dagitty.net (archived at https://perma.cc/U9BY-VZE2) to create DAGs.
Examples of various causal diagrams that could explain why *X* is associated with *Y* are in @fig-XCausesY, @fig-YCausesX, and @fig-YCausesX.

@@

```{r}
#| label: fig-YCausesX
#| fig-cap: "Causal Diagram (Directed Acyclic Graph) Depicting a Third Variable Confound, *Z*, Causing *X* and *Y*, Thus Explaining Why *X* and *Y* are associated."

@@

One of the plausible alternatives to the inference that $X$ causes $Y$ is that there are third variable confounds that influence both $X$ and $Y$, thus explaining why $X$ and $Y$ are associated, as depicted in @fig-correlationAndCausation3 and @fig-YCausesX.
Thus, another approach that can help increase [internal validity](#sec-internalValidity) is to include plausible confounds as control variables.
For instance, if a third variable such as education level might be a confound that influences both sports drink consumption and player performance, you could include education level as a [covariate](#sec-covariates) in the model.
Inclusion of a [covariate](#sec-covariates) attempts to control for the variable by examining the association between the [predictor variable](#sec-correlationalStudy) and the [outcome variable](#sec-correlationalStudy) while holding the [covariate](#sec-covariates) variables constant.
For instance, such a model would examine whether, when accounting for education level, there is an association between sports drink consumption and player performance.

>>>>

A key tool when describing a research question or hypothesis is to create a conceptual depiction of the hypothesized causal processes.
A causal diagram depicts the hypothesized causal processes that link two or more variables.
A common form of causal diagrams is the directed acyclic graph (DAG).
DAGs provide a helpful tool to communicate about causal questions and help identify how to avoid bias (i.e., over-estimation) in associations between variables due to confounding (i.e., common causes) [@Digitale2022].
You can use the `R` package `dagitty` [@Textor2017] or the associated browser-based extension, DAGitty: https://dagitty.net (archived at https://perma.cc/U9BY-VZE2) to create DAGs.
Examples of various causal diagrams that could explain why *X* is associated with *Y* are in @fig-XCausesY, @fig-YCausesX, and @fig-ZCausesXandY.

@@

```{r}
#| label: fig-ZCausesXandY
#| fig-cap: "Causal Diagram (Directed Acyclic Graph) Depicting a Third Variable Confound, *Z*, Causing *X* and *Y*, Thus Explaining Why *X* and *Y* are associated."

@@

One of the plausible alternatives to the inference that $X$ causes $Y$ is that there are third variable confounds that influence both $X$ and $Y$, thus explaining why $X$ and $Y$ are associated, as depicted in @fig-correlationAndCausation3 and @fig-ZCausesXandY.
Thus, another approach that can help increase [internal validity](#sec-internalValidity) is to include plausible confounds as control variables.
For instance, if a third variable such as education level might be a confound that influences both sports drink consumption and player performance, you could include education level as a [covariate](#sec-covariates) in the model.
Inclusion of a [covariate](#sec-covariates) attempts to control for the variable by examining the association between the [predictor variable](#sec-correlationalStudy) and the [outcome variable](#sec-correlationalStudy) while holding the [covariate](#sec-covariates) variables constant.
For instance, such a model would examine whether, when accounting for education level, there is an association between sports drink consumption and player performance.
